---
title: "Using loopQTL"
output:
    BiocStyle::html_document:
        toc: true
        toc_depth: 2
package: loopQTL
author: Rhie Lab at the University of Southern California
description: >
  This vignette describes the basic usage of the loopQTL package to map
  chromatin loop QTLs from Hi-C: which SNPs at a chromatin loop's anchors
  associate with the loop's contact strength across samples.
vignette: >
  %\VignetteIndexEntry{Using loopQTL}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

\RaggedRight

```{r, echo = FALSE, message = FALSE}
## `crop = NULL` resolves "The magick package is required to crop ..." error;
## see: https://stat.ethz.ch/pipermail/bioc-devel/2020-April/016656.html
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", crop = NULL)
options(tibble.print_min = 4, tibble.print_max = 4)
```

## Introduction
loopQTL identifies SNPs associated with chromatin interaction signals across
samples. It implements a band-wise joint normalization approach
(referenced from Brant et al[^Brant]) that systematically tests the
association between variants and their anchored loops, together with downstream
functions for annotation and visualization for the user's convenience.

This vignette walks through the basic loopQTL workflow.

[^Brant] [Fletez-Brant K, Qiu Y, Gorkin DU, Hu M, Hansen KD. Removing unwanted
variation between samples in Hi-C experiments. Brief Bioinform. 2024 Mar
27;25(3):bbae217. doi: 10.1093/bib/bbae217. PMID: 38711367; PMCID:
PMC11074651.](https://doi.org/10.1093/bib/bbae217)

# Acquiring and installing loopQTL and associated packages

## System requirements

R 4.6 or a newer version is required.

On Ubuntu 24.04, successful installation required several additional packages.
They can be installed by running the following command in a terminal:

`sudo apt-get install r-base-dev libcurl4-openssl-dev libfreetype-dev libfribidi-dev libfontconfig-dev libharfbuzz-dev libtiff-dev libxml2-dev libssl-dev`

No dependencies other than R are required on macOS or Windows.

## Installation

Two versions of the package are available.

To install the stable version from Bioconductor, start R and run:

```{r eval = FALSE}
## Install BiocManager, which is required to install packages from Bioconductor
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}

BiocManager::install("loopQTL")
```
The development version containing the most recent updates is available from
our GitHub repository (<https://github.com/rhielab/loopQTL>).

To install the development version from GitHub, start R and run:

```{r eval = FALSE}
## Install prerequisite packages to install the development version from GitHub
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}
if (!requireNamespace("remotes", quietly = TRUE)) {
    install.packages("remotes")
}

BiocManager::install(version = "devel")
BiocManager::install("rhielab/loopQTL")
```

# Overview

1. `loopQTLsetup` - Validate the multi-sample VCF header and per-sample
   `.hic` metadata, read per-sample loop BEDPEs, and build the
   exact-coordinate consensus loop set kept in `loops(obj)`. 
2. `loopQTLnormalize` - Per-chromosome whole-genome straw pull followed by
	band-wise quantile normalization across samples The output is saved in
   `phenotype(obj)`. Optionally caches per-SNP contact-window matrices to a
   file for later visualization with `loopQTLplotSnpWindow()` or
   `loopQTLnormalize()`.
3. `loopQTLchooseK` - Runs PCA on the normalized phenotype matrix and picks
   K via PCAForQTL's method. This function returns the chosen K and the
   corresponding sample PCs.
4. `runQTL` - Tabix-reads anchor-window SNPs from the VCF, runs MatrixEQTL
   to perform the loopQTL analysis. The output is saved in `results(obj)`.
5. `loopQTLpairData` - Return the sample-level phenotype and genotype for a
   single `(loopID, snpID)` pair - the row-level view behind any per-pair
   downstream plot or custom test.
6. `loopQTLannotate` - Adds `gene`, `peak` (optional), and `Type` columns
   to `results(obj)` using a GTF for TSS mapping and an optional peak BED
   file for regulatory-element overlap.
7. `loopQTLplotPair` - Generates boxplot of the loop's contact-strength
   phenotype grouped by the SNP's dosage class (0/1/2), for a specific SNP-loop
   pair.
8. `loopQTLplotLocus` - Locus-level ggplot track combining consensus loops,
   the tested SNP, and (optionally) gene annotations, over a user-specified
   genomic region.
9. `loopQTLplotSnpWindow` - Hi-C-style beta heatmap around a chosen SNP: for
   every bin-pair in the SNP's neighborhood, plots the per-sample
   regression coefficient of phenotype on the SNP's dosage.

To begin, load `loopQTL` and any downstream plotting deps.

```{r setup, message = FALSE}
library(loopQTL)
library(ggplot2)
```

## `loopQTLsetup`: Validate inputs and build loopQTL object

`loopQTLsetup()` is the entry point of the pipeline. It:

* validates the input files header, including .vcf and .hic files;
* reads each per-sample BEDPE and builds a consensus loop set by
  exact-coordinate matching. Loops found fewer than `minSamples` samples will be
  dropped;
* Automatically computes per-sample cis-contact depth from the `.hic` files
  when the samplesheet lacks a `cis` column.

The samplesheet is a tab-separated file with columns `Sample`, `hic`,
`loops`, and optionally `cis`.

```{r}
extdataDir <- system.file("extdata", package = "loopQTL")
vcfPath <- file.path(extdataDir, "example.vcf.gz")
samplesheetPath <- file.path(extdataDir, "example_samplesheet.tsv")
samples <- read.table(samplesheetPath, header = TRUE)

## Fix the input file paths to be absolute since we don't know what the working
## directory will be when the vignette is built. It is not necessary to do this
## during normal use of loopQTL if the working directory is set correctly.
samples$hic <- unlist(
    lapply(samples$hic, function(x) {
        file.path(extdataDir, x)
    })
)
samples$loops <- unlist(
    lapply(samples$loops, function(x) {
        file.path(extdataDir, x)
    })
)

## Display the format of the samplesheet
head(samples)
```

```{r}
## Run loopQTLsetup
obj <- loopQTLsetup(
    samples    = samples,
    vcf        = vcfPath,
    resolution = 10000,
    minSamples = 2
)
obj
```

## `loopQTLnormalize`: Cross-normalization on Hi-C signal

`loopQTLnormalize()` normalizes contact strength across samples so that
per-loop phenotypes are directly comparable. For each chromosome it pulls
the whole-genome straw matrix per sample, log-CPM adjusts, then does
band-wise quantile normalization across samples per genomic-distance band. 
Normalized values at the consensus-loop bin-pairs are written to
`phenotype(obj)`. Parallelizes per chromosome via `BiocParallel`.

Set `saveWindows = TRUE` to additionally output a per-SNP contact-window matrix 
as .rds file. This file is needed by `loopQTLplotSnpWindow()` downstream. Please 
note that this option does not change `phenotype(obj)` or affect the `runQTL()`
step.

```{r}
## Get the path to a temporary directory in which to save the output interact
## file
tempDirectory <- tempdir()

obj <- loopQTLnormalize(
    obj,
    BPPARAM     = BiocParallel::MulticoreParam(workers = 2),
    saveWindows = TRUE,
    windowsOut  = file.path(tempDirectory, "loop_snp_windows.rds"),
    window      = 2000000,
    snpWindow   = 10000
)

head(phenotype(obj)[, 1:5])
```

## `loopQTLchooseK`: Auto-select the number of PCs for covariates

`loopQTLchooseK()` runs PCA on the normalized phenotype matrix and picks a
number of principal components K to use as covariates in the downstream
regression. This function returns a list with the chosen K and the
sample × K matrix of PCs.

```{r}
Kres <- loopQTLchooseK(obj, Kmethod = "elbow")
message("PCAForQTL chose K = ", Kres$K, " via ", Kres$method)
head(Kres$PCs)
```

## `runQTL`: Anchor-restricted association testing

`runQTL()` reads all anchor-window SNPs from the VCF in a single targeted
tabix pass, converts them to dosages (0/1/2), and runs MatrixEQTL twice -
once treating anchor A as the gene position, once anchor B. The two result
tables are outer-joined on SNP-loop level.  Covariates are supplied via the 
`covariates` argument - typically the phenotype PCs chosen in the previous step.
RINT is applied per loop before regression by default.

```{r}
obj <- runQTL(
    obj,
    window     = 10000,
    covariates = Kres$PCs
)
head(results(obj))
```

## `loopQTLpairData`: Sample-level view of one loop × SNP pair

`loopQTLpairData()` returns the per-sample phenotype and genotype
(dosage 0/1/2) for a single SNP-loop pair. Useful for inspecting a
specific hit or exporting sample-level values for reporting.

```{r}
pair <- loopQTLpairData(
    obj,
    loopID = "L0000187_chr1_46030000_46220000",
    snpID  = "rs11584814"
)
head(pair)
```

## `loopQTLannotate`: Annotate results with target genes and regulatory peaks

`loopQTLannotate()` adds three columns to `results(obj)`:

* **`gene`** - comma-separated gene names whose TSS (± `TSSwindow`) sits in
  the SNP's *partner* anchor.
* **`peak`** - logical: does the SNP overlap any peak from the user-supplied
  peak file . If the distal TSS peaks are preferred, then you can set
  `removeTSS = TRUE``.
* **`Type`** - annotation of each SNP-loop pair such as `WithinTSS`, `rs-gene`,
  `rs-<peakAnnotation>`, `rs-O`, `rs(<peakAnnotation>)-gene`, or
  `rs(<peakAnnotation>)-<peakAnnotation>` summarizing the SNP–loop types.

```{r}
peaksPath <- system.file("extdata", "example_annotation.bed", package = "loopQTL")
gtfPath <- system.file("extdata", "example.gtf.gz", package = "loopQTL")
obj <- loopQTLannotate(
    obj,
    peak            = peaksPath,
    peakAnnotation  = "Enhancer",
    TSSwindow       = 2000,
    removeTSS       = TRUE,
    gtf             = gtfPath
)
head(results(obj))
```

## `loopQTLplotPair`: Boxplot of loop strength by SNP dosage

`loopQTLplotPair()` renders the per-sample boxplot behind a chosen result
SNP-loop pair. Please note that the plot is a ggplot object that you can save
with `ggsave()`.

```{r}
p <- loopQTLplotPair(
    obj,
    loopID = "L0000187_chr1_46030000_46220000",
    snpID  = "rs11584814"
)
p
```

## `loopQTLplotLocus`: Locus-level track view of loops, SNPs, and genes

`loopQTLplotLocus()` renders a locus-scale ggplot combining the consensus
loops, the tested SNP, and (when a GTF is supplied) gene annotations, over
a user-specified region. Please note that to sse `values = "cpm"` or 
`values = "raw"` to show per-loop contact strength, you need to set
`saveWindows = TRUE`  to save the per-SNP contact-window matrices from
loopQTLnormalize.

```{r}
gtfPath <- system.file("extdata", "example.gtf.gz", package = "loopQTL")
p <- loopQTLplotLocus(
    obj,
    chrom = "chr1",
    start = 45900000,
    end = 46300000,
    snpID = "rs11584814",
    values = "cpm",
    gtf = gtfPath
)
p
```

## `loopQTLplotSnpWindow`: Hi-C-style beta heatmap around a SNP

`loopQTLplotSnpWindow()` renders a Hi-C-style heatmap centered on a chosen
SNP: for every bin-pair in the SNP's cached contact window (from
`loopQTLnormalize(saveWindows = TRUE)`), the color encodes the per-sample
regression coefficient (beta) of the bin-pair's contact strength on the SNP's
dosage. Set `normal = TRUE` and `covariates = "auto"` to match the model
used in `runQTL()`.

```{r}
p <- loopQTLplotSnpWindow(
    obj,
    chrom      = "chr1",
    start      = 45900000,
    end        = 46300000,
    snpID      = "rs11584814",
    normal     = TRUE,
    covariates = "auto",
    Kmax       = 2
)

## Delete the output file, since this is just an example.
## Do not use this line if running on real data, as it will delete your created
## file.
invisible(file.remove(metadata(obj)$windowsOut))

p
```

# Session info

```{r}
sessionInfo()
```
